Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-26 article/main.tex
Small fixes and comments marked as MUN1

The word \textit{startup} appeared in the SE literature for the first time in 
1994 in an article written by Carmel \cite{Camel1994a} where he studied the 
time-to-completion in a young package firm. %MUN: If this is a quote 
%from Carmel, put it in quotation marks, like you did in the previous section 
%with Suttons quote. Generally, check that you %are consistent how you use 
%quotations throughout the paper. 
%CG: fixed, btw  it was just to emphasize his the focus of his study. 
He noticed how these companies were particularly innovative and successful, 
advocating for the need of more research investigating software development 
practices so as to replicate success and try to transfer it to other technology 
sectors.

@@

Additionally, the role of SPI has been neglected because seen as an obstacle to 
the development team's creativity and flexibility, and to 
the need of a quick delivering product process environment \cite{Coleman2008a}.%MUN: the second part of the sentence is just a statement without reference. If it refers to a 
%statement by Coleman, put his reference at the end OR find the correct 
%reference OR delete the statement.
%CG fixed
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies of Mater and Mirel 
\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in 
internet startups, are oriented to usability and scalability, even though the 
market and application type heavily influences the quality-demand 
\cite{Coleman2008, Kim2005}.

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=5in]{figures/completemethodology2}
\caption{Grounded theory study process overview}
\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby
%to keep the height of the figure, expanding however the width (use horizontal
%space since it is a two column figure). Change also in figure:  - Write "Data
%Analysis" horizontally  - Change names of macro-phases to upper-case, i.e.
%Design and Execution,     Data Collection, Data Analysis, etc.  - Add section
%number to macro-phase name, i.e. 3.1 Design and Execution, 3.2     Data
%Collection  - Change "Discussion and Validation" into "Theory Validation" as
%you name it     in the text. Also, add a section (3.5) that describes shortly
%what you did in "Theory  Validation" CG: fixed... fontsize cannot be increased
%uniformly among all the boxes, then I skipped that part...

@@

All companies, except C10, were founded within the last three years (2009-2012), 
%MUN: I've shortened the text and added absolute numbers. Is 2009-2012 correct? CG: right
by an average of 3 starting members, which were in majority developers. 
Moreover, the number of current employees shows how, to different degrees, 
companies expanded the initial teams. All companies, except C5, brought their 
first product to market in less than 6 months from the idea conception. The 
products consist of web-applications, launched in different nations and market 
sectors. The growing team size and the publicly available data suggest an 
healthy status of the businesses.

@@

The follow-up questionnaires were designed to capture additional data, gather 
missing information and confirm interview results by triangulation. The 
questionnaires have been tailored to each startup, partially taking advantage of 
the repertory grid principles \cite{Edwards2009}. To obtain better chances of 
quickly obtaining responses, questionnaires have been sent to respondents 
immediately after transcribing the interview results. Most of the follow-up questionnaires have been collected within one day after the 
conclusion of the last interview.   %MUN: I don't understand this: you
%collected the data from questionnaires after ALL interviews have been 
%finished, but on the other hand you said the overall process was iterative 
%(data analysis and triangulation let themes emerge which in turn were used
%in subsequent interviews). CG: fixed...

@@

%------------------------------------------------------------------------------%
\subsection{Theory Validation}
\label{rm:val}
%MUN: put here what I wrote in the comment in Section "Theory validation" (i.e.
%the section following the GSM)
%CG: fixed

Presenting GT is challenging for a researcher, who must pay attention to structure the level of details included, and to the way data is portrayed to display evidence of the emergent categories.  In order to assess our research and determine if the GT is sufficiently grounded, we used a systematic technique to assess the theory, following evaluation lists and factors presented in  \cite{Strauss1998}.

First we had to consider the following factors:

\par\vspace {0.4cm}
\begin{compactitem}
\item \textit{Fit}, that describes the consistence of the theory with the real data.
\item \textit{Understanding}, that describes the degree to which participants agree with the theory.
\item \textit{Generality}, that describes the level of abstraction of the theory, and how sufficiently it can be adopted by practitioners without losing its relevance.
\item \textit{Control}, that describes how well the theory enables practitioners to fully understand the described situation.
\end{compactitem}
\par\vspace {0.4cm}

@@

\par\vspace {0.4cm}
\begin{compactenum}
\item Are concepts generated, and are the concepts systematically related?
\item Are there many conceptual linkages and are the categories well developed?
\item Is variation\footnote{Variation refers to the variety of contexts which theory can be applied.} built into the theory and are the conditions under which variation can be found built into the study and explained?
\item Are the conditions under which variation can be found built into the study and explained?
\item Has the process been taken into account, and does the theory stand the test of time?
\item Do the theoretical findings seem significant, and to what extent?
\end{compactenum}
\par\vspace {0.4cm}

Answering the above questions, the concepts were generated according to the coding process. They were systematically related through the use of a network diagram (see figure \ref{fig:gt:completemethodology}), by which we established the linkages and relations between the concepts. At each iteration of the grounded theory process, we considered and examined a concept within different conditions and dimensions, trying to incorporate opinions of a broader as possible range of practitioners.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical 
categories using the \textit{italic} font style. For the main categories shown 
in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. 
CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up 
development}, which is the most interconnected node in the framework reflecting 
the fact that \textit{``it is the one (category) with the greatest explanatory 
power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that 
reach the core category (efficiency, effectiveness and performance) is discussed 
in detail in section \ref{res:val}. %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?
%CG:fixed

@@

The \textit{severe lack of resources} forces the company to focus on 
implementing an essential set of functionalities and it is one of the main 
reasons why the \textit{product quality has low priority} respect to other more 
urgent needs\footnote{There are some exceptions where the quality aspects 
actually matter and such cases will be discussed in section \ref{res:val}%MUN: 
%this refers to the validity threats section. Do you mean rather the "Theory 
%Validation" section?
%CG fixed
.}. At the same time, to be able to deal with such constraints, startups must 
depend on a small group of capable and motivated individuals.

@@

A \textit{limited access to expertise} forces the team to rely mainly on their 
personal abilities, even though asking opinions to mentors is reported to be a 
viable practice to maintain feasible objectives. Furthermore \textit{teams work 
under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to 
%this statement since it is a citation? 
%CG: not fixed. It represents a subcategory, expressed by the 70% of the interviewed startups
mainly constrained by a tight \textit{time shortage}.

@@

To achieve a good level of UX while dealing with lack of human resources and 
time shortages, startups can analyze similar products of bigger companies which 
can afford more rigorous usability studies. Then, the users' feedback and 
product metrics begin to have a central role in determining the level of UX 
achieved. Product metrics are typically web-based \textit{statistical hypothesis 
testing}, such as A/B testing \cite{AB}. %MUN: add reference or explain shortly what A/B 
%testing is
%CG: fixed

@@

Furthermore, a rigid project management is perceived as \textit{``waste of 
time''} which hinders the development speed since the \textit{uncertainty makes 
formal scheduling pointless} (C9  reports that \textit{``initial chaos helps to 
develop faster''}). Startups' \textit{minimal project management}  is supported 
by keeping: \textit{internal milestones short and informal}, 
\textit{low-precision task assignment mechanism}  and a very low attention for 
project metrics (paraphrasing C13, \textit{``the only track of progress was made 
by looking at closed tickets''}). In this context \textit{only a final release 
milestone is viable}, which helps practitioners to remain focused on short term 
goals and put new features in production. To further \textit{speed-up 
development}, startups \textit{simplify issue management by integrating it with 
feature management}. Then, the \textit{limited number of functionalities} and 
the {use of standard/known technologies} with a \textit{simple workflow} are the 
main reasons for 
not establishing a heavy project management process since the project and 
technologies can be autonomously managed by the development team (with 
characteristics of: co-location, self-organization and very small team size). %MUN: 
%do you mean small team size?
%CG: fixed

@@

In this section we discuss the validity of the GSM by means of 
cross-methodological observations, as discussed in section \ref{rm:val}.
%MUN: expand this discussion a bit (motivation why you do this like you do it) 
%and move to Section 3.5. Then write here just that you perform theory 
%validation as discussed in Section 3.5 (and remove the bullet list below).
%CG:fixed

@@

Coleman's framework aims to highlight how managers consider two distinct kinds 
of processes: \textit{essentials} and \textit{non-essentials}. The 
\textit{essential} processes are the most closely linked to product development, 
such as requirements gathering, design and testing. The \textit{non-essential} 
processes are those that might be omitted, such as planning, estimating and 
staging meetings. In particular, he discusses how practices are routinely 
removed: \textit{``With most methodologies and approaches, very few stick to the 
letter of them and they are always adapted, so we adapted ours to the way we 
wanted it to work for us, for our own size and scale''} \cite{Coleman2008}. %MUN: add ref to 
%Coleman's paper where he states this
%CG fixed

@@

In conclusion, by the comparison of the GSM with Coleman's framework, we have
found that despite Coleman's study contributes to define which factors hinder
the introduction of SPI in startups %MUN: add: "of SPI in startups"? CG: fixed ,
his results are coherent with the theory generated within our research, focusing
at explaining the overall strategies within software development in
startups%MUN:  %add "our research, focused at explaining software development in
general in  %startups." ? CG:fixed . In this regard, a wide variability of
processes is a key factor for startups.  In contrast with some SPI processes,
such as \textit{Six Sigma} which objective  is minimizing unpredictability in
the definition of workflows \cite{Sixsigma},  startups seem to follow the
\textit{Fischer's Fundamental Theorem of natural  selection} \cite{Fisher}.
Moving towards flexible and variable processes (easy  to adapt to the uncertain
conditions), increase the odds of ``natural  selection'' only when some
``restrictive conditions'' are met. In other words,  even though ignoring SPI
practices is advisable, under circumstances of high uncertainty and severe lack
of resources, %MUN: under which  %circumstances? This statement sounds too
general, i.e. don't do SPI ever... %CG: fixed %And the following part,I did not
fully understand. Try to clarify. CG:fixed  not following SPI does not predict
major adaptiveness of startups development  towards the customer's needs.
Nevertheless, to imply a major evolutionary product/market fit, certain
conditions explained in GSM need to be considered.

@@

%CG: partially fixed. Agreed on the beginning of the section, not agreed on the division in subsections. Let's discuss it by a skype call.

Also Baskerville \cite{Internet} refers to SPI as a development approach
typically effective only in large-scale, long-term development efforts with
stable and disciplined processes.Internet-speed software development (i.e. an
agile method oriented using daily builds aimed at developing a product with high
speed) differs from traditional software development. Baskerville, by applying a
grounded theory analysis in 10 companies, found that major causal factors that
influence development are:

@@

Only 7 studies out of 37, discuss aspects related to the overall software 
development process in early stage startups. The remaining part of the studies 
is only partially covering the phenomenon described by the GSM. Looking at the 
papers' main contribution %MUN: which results? The papers' main contribution or...? %CG: fixed
, we were able to extrapolate information which confirmed the GSM categories.

@@

%MUN: Think of structuring this long subsection. Have a subsubsection for each 
%CAT, i.e. 5.2.1 Severe lack of resources (CAT7)
%I've added to each category you mention the corresponding CATx since that  
%identifier you use in table 2. I did not check if you discuss each and every 
%category. Do you? If yes, ok. If not, why not? CG: partially fixed, the categories CAT3 and CAT6 are not well covered as discussed above.
Starting from the category \textit{severe lack of resources} (CAT7), many 
authors describe startups facing with inescapable constraints defined by 
engineering/business concerns and constant time-pressure \cite{Camel1994a, 
Cugola98softwareprocesses:}. Because of their young and inexperienced working 
conditions, startups present limited resources in terms of management and 
strategic alliances \cite{Sutton2000}.

Crowne \cite{Crowne2002} identifies four main stages of a startup's lifecycle: 
startup, stabilization, growth and maturity. Inexperienced practitioners 
operating in a chaotic environment characterizes the first phase, where there is 
no strategic plan for developing the product. But, moving towards the 
stabilization phase, the product starts to be unreliable and requirements 
unmanageable. Consequently, when startups move to the growth phase, they 
initiate structuring and controlling processes, which are incrementally 
integrated in the development environment. The maturity phase represents the 
moment when the company is ready to introduce process improvement initiatives. 
Crowne's study profoundly corresponds to different conceptualizations in the 
GSM. Starting from the minimal and low-precision engineering elements, startups 
start to grow the accumulated technical debt which requires fast integration of 
more structured and standard processes overtime.
Issues related to low-precision engineering practices during companies' growth 
are discussed in an experience paper by Ambler \cite{Ambler2002}. He reports on 
two case studies of growing Internet startups approaching an initial public offering (IPO). %MUN:explain 
%acronym %CG: fixed
Despite the two companies were facing business and team growth, attitudes 
against the introduction of processes to structure and model the system were 
evident. The need to re-engineer the codebase lead the researchers to suggest a 
tailored version of the Rational Unified Process (RUP) %MUN: add reference
 %CG: not fixed, not sure if needed since Ambler has the detailed 
 %explaination of his process, similar to RUP... anyway a dedicated resource would be an IBM website
development framework. Nevertheless, even with the aid of CASE 
tools\footnote{CASE tools are a class of software that automates many of the 
activities involved in various engineering activities. See 
\url{http://www.unl.csi.cuny.edu/faqs/software-enginering/tools.html} .}, 
startups failed in practicing engineering activities in view of excessive lack of 
flexibility. Developers fast moved to simpler tools such as using a white-board, 
where they were able to provide input and insights, question assumptions, share 
past experiences, and even talk about potential implementation strategies. 
%MUN: Candidate for removal since details can be found in Ambler anyway: "In 
%general the development remained highly iterative, where project teams started 
%to have a little modeling and testing only when necessary. They didn't take 
%the serial approach of completely developing and accepting the requirements 
%model, analysis model and so on. This let them react swiftly and adapt 
%rapidly according to changes into their highly competitive markets. The 
%researchers report \textit{``modeling was streamlined''}, that is modeling 
%what was needed and go straightforward to coding, avoiding analysis 
%paralysis." %CG: ok

@@

%MUN: the whole paragraph needs to be summarized to its essence. Details can be 
%read in the referred paper. CG: partially fixed, removed the detailed process description
In the evaluation phase, the status of the development workflow within the 
company  is analyzed. The researchers documented a state of chaos where any 
initiatives of quantitative methodology assessment would have been a waste of 
time. Therefore, they applied a qualitative methodology to define a model of the 
release management process. 
%They moved towards the plan phase where the 
%researchers decided to develop a model to manage releases, introducing a process 
%composed by: release scope preparation, release planning, release development, 
%system testing, acceptance testing and release deployment. 
Despite the effort 
spent by the researchers to establish a development workflow, practices were not 
fully conducted even though some marginal benefits have been reported: planned 
major release from four to six months; planned minor releases from two to four 
weeks; encompassing urgent corrective releases from one to five days. Regarding 
the previously mentioned problems, only 
partial improvements have been achieved since the major obstacle was to 
motivate employees to change their habits. With perfect hindsight the 
researchers claim that understanding and adapting solutions by solving one 
problem at a time would have increased the benefits. Then, despite the reported 
results, starting the establishment of a simple and informal workflow aided the 
organization to move from the stage of knowing nothing to knowing at least 
something about their development process.

@@

Prescribed development models, oriented to SPI, are basically ignored by 
startups, as discussed in \cite{Zettel2001}. %MUN: <-- rephrase, not clear to me %CG:fixed
To overcome the need of structure and to accelerate the development process, 
Zettel developed a \textit{``LIghtweight Process for E-business software 
development''} (LIPE). This lightweight development method integrates the XP 
approach with ideas from areas of software measurement for project control and 
process improvement as a compromise between an ad-hoc and a more rigorous 
approach. Zettel argues that LIPE provides scalability of the development 
process with a certain degree of flexibility, omitting parts of prescribed 
practices. Nevertheless, we couldn't identify any empirical evaluation of LIPE 
that supports its suitability in a startup context.

Deakins \cite{Deakins2005} designed a new spiral methodology, constructed during a
fieldwork with a developer team that had limited resources managing an 
innovative product in volatile e-commerce environments. This methodology represents a fast 
evolutionary approach experimentation with flexible processes. It aims to obtain
customer's feedback as soon as possible for continuous product improvement.%MUN: 
%<-- rephrase, sentence not parsable CG:fixed
Even in this case the spiral methodology%MUN: which model? CG:fixed
is compliant with the nature of a random, ad-hoc and \textit{evolutionary development approach} (CAT2). In fact, evolutionary development approaches are typically more 
suitable  for web applications, as confirmed in \cite{Deshpande2001}.
Moreover, they provide practices oriented towards rapid and iterative design with complete freedom 
to the developer team, enabled to operate anywhere within the organization (compliant also to the category.%MUN: <-- the whole paragraph is not that clear to me. Try to start 
%with the category (team is catalyst), then write about Deakins spiral model
%CG: partially fixed, the focus here is on the evolutionary development approach

>>>>

The word \textit{startup} appeared in the SE literature for the first time in 
1994 in an article written by Carmel \cite{Camel1994a} where he studied the 
time-to-completion in a young package firm. %MUN: If this is a quote 
%from Carmel, put it in quotation marks, like you did in the previous section 
%with Suttons quote. Generally, check that you are consistent how you use 
%quotations throughout the paper. 
%CG: fixed, btw  it was just to emphasize his the focus of his study. 
%MUN1: Ok. As you may have noticed, I have removed a lot of cases were you use 
%italics. If you emphasize *too* much, it it rather distracting than helpful 
%for the reader. Furthermore, it conflicts with you convention you use to 
%highlight high level categories of GSM.
He noticed how these companies were particularly innovative and successful, 
advocating for the need of more research investigating software development 
practices so as to replicate success and try to transfer it to other technology 
sectors.

@@

Additionally, the role of SPI has been neglected because seen as an obstacle to 
the development team's creativity and flexibility, and to 
the need of a quick delivering product process environment \cite{Coleman2008a}.
%MUN: the second part of the sentence is just a statement without reference. If 
%it refers to a statement by Coleman, put his reference at the end OR find the 
%correct reference OR delete the statement.
%CG fixed
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies of Mater and Mirel 
\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in 
internet startups, are oriented to usability and scalability, even though the 
market and application type heavily influences the quality-demand 
\cite{Coleman2008, Kim2005}.

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=5in]{figures/completemethodology2}
\caption{Grounded theory study process overview}
\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby
%to keep the height of the figure, expanding however the width (use horizontal
%space since it is a two column figure). Change also in figure:  - Write "Data
%Analysis" horizontally  - Change names of macro-phases to upper-case, i.e.
%Design and Execution,     Data Collection, Data Analysis, etc.  - Add section
%number to macro-phase name, i.e. 3.1 Design and Execution, 3.2     Data
%Collection  - Change "Discussion and Validation" into "Theory Validation" as
%you name it     in the text. Also, add a section (3.5) that describes shortly
%what you did in "Theory  Validation" 
%CG: fixed... fontsize cannot be increased uniformly among all the boxes, then 
%I skipped that part...
%MUN1: OK, but in the figure, the correct section number for theory validation 
%is 3.5, not 3.4

@@

All companies, except C10, were founded within the last three years (2009-2012), 
%MUN: I've shortened the text and added absolute numbers. Is 2009-2012 correct? 
%CG: right
by an average of 3 starting members, which were in majority developers. 
Moreover, the number of current employees shows how, to different degrees, 
companies expanded the initial teams. All companies, except C5, brought their 
first product to market in less than 6 months from the idea conception. The 
products consist of web-applications, launched in different nations and market 
sectors. The growing team size and the publicly available data suggest an 
healthy status of the businesses.

@@

The follow-up questionnaires were designed to capture additional data, gather 
missing information and confirm interview results by triangulation. The 
questionnaires have been tailored to each startup, partially taking advantage of 
the repertory grid principles \cite{Edwards2009}. To obtain better chances of 
quickly obtaining responses, questionnaires have been sent to respondents 
immediately after transcribing the interview results. Most of the follow-up  
questionnaires have been collected within one day after the conclusion of the 
last interview.   
%MUN: I don't understand this: you
%collected the data from questionnaires after ALL interviews have been 
%finished, but on the other hand you said the overall process was iterative 
%(data analysis and triangulation let themes emerge which in turn were used
%in subsequent interviews). 
%CG: fixed...

@@

%------------------------------------------------------------------------------%
\subsection{Theory Validation}
\label{rm:val}
%MUN: put here what I wrote in the comment in Section "Theory validation" (i.e.
%the section following the GSM)
%CG: fixed
%MUN1: This description is way too long and verbose. It should be MAX 1 column, 
%and focus on what you present in Section 5. Theory Validation. Until the 
%second-last paragraph you don't talk about that at all! Remove the "theory" on 
%GT validation and what you did *during* the GT process. Keep what you need 
%to motivate
% - 5.1 Comparison of frameworks
% - 5.2 Theoretical categories and existing literature
% - 5.3 Confounding factors
% - 5.4 Static assessment of the GSM
%You need to explain why you did this and how it contributes to the validation 
%of the theory.

Presenting GT is challenging for a researcher, who must pay attention to  
structure the included level of detail, and to the way data is portrayed to 
display evidence of the emergent categories.  In order to assess our research 
and determine if the GT is sufficiently grounded, we used a systematic technique 
to assess the theory, following evaluation lists and factors presented in  
\cite{Strauss1998}.

We consider the following factors:

\par\vspace {0.4cm}
\begin{compactitem}
\item \textit{Fit}, that describes the consistency of the theory with the real 
data.
\item \textit{Understanding}, that describes the degree to which participants 
agree with the theory.
\item \textit{Generality}, that describes the level of abstraction of the 
theory, and how sufficiently it can be adopted by practitioners without losing 
its relevance.
\item \textit{Control}, that describes how well the theory enables 
practitioners to fully understand the described situation.
\end{compactitem}
\par\vspace {0.4cm}

@@

\par\vspace {0.4cm}
\begin{compactenum}
\item Are concepts generated, and are the concepts systematically related?
\item Are there many conceptual linkages and are the categories well developed?
\item Is variation\footnote{Variation refers to the variety of contexts to 
which the theory can be applied.} built into the theory and are the conditions 
under which variation can be found built into the study and explained?
\item Are the conditions under which variation can be found built into the 
study and explained?
\item Has the process been taken into account, and does the theory stand the 
test of time?
\item Do the theoretical findings seem significant, and to what extent?
\end{compactenum}
\par\vspace {0.4cm}

Answering the above questions, the concepts were generated according to the 
coding process. They were systematically related through the use of a network 
diagram (see Figure \ref{fig:gt:completemethodology}), by which we established 
the relations between the concepts. At each iteration of the 
grounded theory process, we considered and examined a concept within different 
conditions and dimensions, trying to incorporate data from a broader range of 
practitioners.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical 
categories using the \textit{italic} font style. For the main categories shown 
in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. 
CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up 
development}, which is the most interconnected node in the framework reflecting 
the fact that \textit{``it is the one (category) with the greatest explanatory 
power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that 
reach the core category (efficiency, effectiveness and performance) is discussed 
in detail in section \ref{res:val}. %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?
%CG:fixed
%MUN1: this should be \ref{sect:theory:validation:stats}, right? Section 5 is 
%quite long and you should point to the correct subsection

@@

The \textit{severe lack of resources} forces the company to focus on 
implementing an essential set of functionalities and it is one of the main 
reasons why the \textit{product quality has low priority} respect to other more 
urgent needs\footnote{There are some exceptions where the quality aspects 
actually matter and such cases will be discussed in section \ref{res:val}%MUN: 
%this refers to the validity threats section. Do you mean rather the "Theory 
%Validation" section?
%CG fixed
%MUN1: Same here, point to the correct subsection.
.}. At the same time, to be able to deal with such constraints, startups must 
depend on a small group of capable and motivated individuals.

@@

A \textit{limited access to expertise} forces the team to rely mainly on their 
personal abilities, even though asking opinions to mentors is reported to be a 
viable practice to maintain feasible objectives. Furthermore \textit{teams work 
under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to 
%this statement since it is a citation? 
%CG: not fixed. It represents a subcategory, expressed by the 70% of the 
%interviewed startups
%MUN1: and how should the reader know that? See my problem with italics: you 
%use italics for both interview statements and (sub-)categories. 
mainly constrained by a tight \textit{time shortage}.

@@

To achieve a good level of UX while dealing with lack of human resources and 
time shortages, startups can analyze similar products of bigger companies which 
can afford more rigorous usability studies. Then, the users' feedback and 
product metrics begin to have a central role in determining the level of UX 
achieved. Product metrics are typically web-based \textit{statistical hypothesis 
testing}, such as A/B testing \cite{AB}. 
%MUN: add reference or explain shortly what A/B testing is
%CG: fixed

@@

Furthermore, a rigid project management is perceived as \textit{``waste of 
time''} which hinders the development speed since the \textit{uncertainty makes 
formal scheduling pointless} (C9  reports that \textit{``initial chaos helps to 
develop faster''}). Startups' \textit{minimal project management}  is supported 
by keeping: \textit{internal milestones short and informal}, 
\textit{low-precision task assignment mechanism}  and a very low attention for 
project metrics (paraphrasing C13, \textit{``the only track of progress was made 
by looking at closed tickets''}). In this context \textit{only a final release 
milestone is viable}, which helps practitioners to remain focused on short term 
goals and put new features in production. To further \textit{speed-up 
development}, startups \textit{simplify issue management by integrating it with 
feature management}. Then, the \textit{limited number of functionalities} and 
the {use of standard/known technologies} with a \textit{simple workflow} are the 
main reasons for 
not establishing a heavy project management process since the project and 
technologies can be autonomously managed by the development team (with 
characteristics of: co-location, self-organization and very small team size). 
%MUN: do you mean small team size?
%CG: fixed

@@

In this section we discuss the validity of the GSM by means of 
cross-methodological observations, as discussed in section \ref{rm:val}.
%MUN: expand this discussion a bit (motivation why you do this like you do it) 
%and move to Section 3.5. Then write here just that you perform theory 
%validation as discussed in Section 3.5 (and remove the bullet list below).
%CG:fixed
%MUN1: Ok, but fix section 3.5... you see that you don't motivate 5.1-5.4 at 
%all there and that the current content of 3.5 is not related to what you 
%actually do here as validation?

@@

Coleman's framework aims to highlight how managers consider two distinct kinds 
of processes: \textit{essentials} and \textit{non-essentials}. The 
\textit{essential} processes are the most closely linked to product development, 
such as requirements gathering, design and testing. The \textit{non-essential} 
processes are those that might be omitted, such as planning, estimating and 
staging meetings. In particular, he discusses how practices are routinely 
removed: \textit{``With most methodologies and approaches, very few stick to the 
letter of them and they are always adapted, so we adapted ours to the way we 
wanted it to work for us, for our own size and scale''} \cite{Coleman2008}. 
%MUN: add ref to Coleman's paper where he states this
%CG fixed

@@

In conclusion, by the comparison of the GSM with Coleman's framework, we have
found that despite Coleman's study contributes to define which factors hinder
the introduction of SPI in startups 
%MUN: add: "of SPI in startups"? CG: fixed
his results are coherent with the theory generated within our research, focusing
at explaining the overall strategies within software development in startups.
%MUN: add "our research, focused at explaining software development in
%general in startups." ? CG:fixed
In this regard, a wide variability of processes is a key factor for startups.  
In contrast with some SPI processes, such as \textit{Six Sigma} which objective 
is minimizing unpredictability in the definition of workflows \cite{Sixsigma}, 
startups seem to follow the \textit{Fischer's Fundamental Theorem of natural  
selection} \cite{Fisher}.
Moving towards flexible and variable processes (easy  to adapt to the uncertain
conditions), increase the odds of ``natural  selection'' only when some
``restrictive conditions'' are met. In other words, even though ignoring SPI
practices is advisable, under circumstances of high uncertainty and severe lack
of resources, %MUN: under which  %circumstances? This statement sounds too
%general, i.e. don't do SPI ever... %CG: fixed 
%And the following part,I did not fully understand. Try to clarify. CG:fixed  
not following SPI does not predict
major adaptiveness of startups development towards the customer's needs.
Nevertheless, to imply a major evolutionary product/market fit, certain
conditions explained in GSM need to be considered.

@@

%CG: partially fixed. Agreed on the beginning of the section, not agreed on  the 
%division in subsections. Let's discuss it by a skype call.

Also Baskerville \cite{Internet} refers to SPI as a development approach
typically effective only in large-scale, long-term development efforts with
stable and disciplined processes. Internet-speed software development (i.e. an
agile method oriented using daily builds aimed at developing a product with high
speed) differs from traditional software development. Baskerville, by applying a
grounded theory analysis in 10 companies, found that major causal factors that
influence development are:

@@

Only 7 studies out of 37, discuss aspects related to the overall software 
development process in early stage startups. The remaining part of the studies 
is only partially covering the phenomenon described by the GSM. Looking at the 
papers' main contribution%MUN: which results? The papers' main contribution 
%or...? %CG: fixed
, we were able to extrapolate information which confirmed the GSM categories.

@@

%MUN: Think of structuring this long subsection. Have a subsubsection for each 
%CAT, i.e. 5.2.1 Severe lack of resources (CAT7)
%I've added to each category you mention the corresponding CATx since that  
%identifier you use in table 2. I did not check if you discuss each and every 
%category. Do you? If yes, ok. If not, why not? 
%CG: partially fixed, the categories CAT3 and CAT6 are not well covered as 
%discussed above.
Starting from the category \textit{severe lack of resources} (CAT7), many 
authors describe startups facing with inescapable constraints defined by 
engineering/business concerns and constant time-pressure \cite{Camel1994a, 
Cugola98softwareprocesses:}. Because of their young and inexperienced working 
conditions, startups present limited resources in terms of management and 
strategic alliances \cite{Sutton2000}.

Crowne \cite{Crowne2002} identifies four main stages of a startup's lifecycle: 
startup, stabilization, growth and maturity. Inexperienced practitioners 
operating in a chaotic environment characterizes the first phase, where there is 
no strategic plan for developing the product. But, moving towards the 
stabilization phase, the product starts to be unreliable and requirements 
unmanageable. Consequently, when startups move to the growth phase, they 
initiate structuring and controlling processes, which are incrementally 
integrated in the development environment. The maturity phase represents the 
moment when the company is ready to introduce process improvement initiatives. 
Crowne's study profoundly corresponds to different conceptualizations in the 
GSM. Starting from the minimal and low-precision engineering elements, startups 
start to grow the accumulated technical debt which requires fast integration of 
more structured and standard processes overtime.
Issues related to low-precision engineering practices during companies' growth 
are discussed in an experience paper by Ambler \cite{Ambler2002}. He reports on 
two case studies of growing Internet startups approaching an initial public  
offering (IPO). %MUN:explain %acronym %CG: fixed
Despite the two companies were facing business and team growth, attitudes 
against the introduction of processes to structure and model the system were 
evident. The need to re-engineer the codebase lead the researchers to suggest a 
tailored version of the Rational Unified Process (RUP) 
%MUN: add reference
%CG: not fixed, not sure if needed since Ambler has the detailed 
%explaination of his process, similar to RUP... anyway a dedicated resource 
%would be an IBM website
development framework. Nevertheless, even with the aid of CASE 
tools\footnote{CASE tools are a class of software that automates many of the 
activities involved in various engineering activities. See 
\url{http://www.unl.csi.cuny.edu/faqs/software-enginering/tools.html} .}, 
startups failed in practicing engineering activities in view of excessive lack of 
flexibility. Developers fast moved to simpler tools such as using a white-board, 
where they were able to provide input and insights, question assumptions, share 
past experiences, and even talk about potential implementation strategies. 
%MUN: Candidate for removal since details can be found in Ambler anyway: "In 
%general the development remained highly iterative, where project teams started 
%to have a little modeling and testing only when necessary. They didn't take 
%the serial approach of completely developing and accepting the requirements 
%model, analysis model and so on. This let them react swiftly and adapt 
%rapidly according to changes into their highly competitive markets. The 
%researchers report \textit{``modeling was streamlined''}, that is modeling 
%what was needed and go straightforward to coding, avoiding analysis 
%paralysis." %CG: ok

@@

%MUN: the whole paragraph needs to be summarized to its essence. Details can be 
%read in the referred paper. 
%CG: partially fixed, removed the detailed process description
In the evaluation phase, the status of the development workflow within the 
company  is analyzed. The researchers documented a state of chaos where any 
initiatives of quantitative methodology assessment would have been a waste of 
time. Therefore, they applied a qualitative methodology to define a model of the 
release management process. 
%They moved towards the plan phase where the 
%researchers decided to develop a model to manage releases, introducing a process 
%composed by: release scope preparation, release planning, release development, 
%system testing, acceptance testing and release deployment. 
Despite the effort spent by the researchers to establish a development workflow, 
practices were not fully conducted even though some marginal benefits have been 
reported: planned major release from four to six months; planned minor releases 
from two to four weeks; encompassing urgent corrective releases from one to five 
days. 
Regarding the previously mentioned problems, only 
partial improvements have been achieved since the major obstacle was to 
motivate employees to change their habits. With perfect hindsight the 
researchers claim that understanding and adapting solutions by solving one 
problem at a time would have increased the benefits. Then, despite the reported 
results, starting the establishment of a simple and informal workflow aided the 
organization to move from the stage of knowing nothing to knowing at least 
something about their development process.

@@

Prescribed development models, oriented towards SPI, are basically ignored by 
startups, as discussed in \cite{Zettel2001}. 
%MUN: <-- rephrase, not clear to me %CG:fixed
To overcome the need of structure and to accelerate the development process, 
Zettel developed a \textit{``LIghtweight Process for E-business software 
development''} (LIPE). This lightweight development method integrates the XP 
approach with ideas from areas of software measurement for project control and 
process improvement as a compromise between an ad-hoc and a more rigorous 
approach. Zettel argues that LIPE provides scalability of the development 
process with a certain degree of flexibility, omitting parts of prescribed 
practices. Nevertheless, we couldn't identify any empirical evaluation of LIPE 
that supports its suitability in a startup context.

Deakins \cite{Deakins2005} designed a new spiral methodology, constructed  
during a fieldwork with a developer team that had limited resources managing an 
innovative product in volatile e-commerce environment. This methodology 
represents a fast evolutionary approach with flexible processes. 
It aims to obtain customer's feedback as soon as possible for continuous 
product improvement.%MUN: %<-- rephrase, sentence not parsable CG:fixed
Even in this case, the spiral methodology%MUN: which model? CG:fixed
is compliant with the nature of a random, ad-hoc and  \textit{evolutionary 
development approach} (CAT2). Evolutionary development approaches are 
typically more suitable for web applications, as confirmed in 
\cite{Deshpande2001}.
Moreover, they provide practices oriented towards rapid and iterative design 
with complete freedom to the developer team, enabled to operate anywhere within 
the organization.
%MUN: <-- the whole paragraph is not that clear to me. Try to start 
%with the category (team is catalyst), then write about Deakins spiral model
%CG: partially fixed, the focus here is on the evolutionary development approach
